The Role of Advection Versus Diffusion in Intraventricular Transport

### 2025-02-24 paper/main.tex
discussion and appendix edits, update bib

\subsection{Computational model of solute transport after photoconversion}
We simulate transport of photoconverted proteins within the brain ventricles by
modeling transport of a solute with concentration $c(\xx, t)$ for $\xx\in\Omega$ and
time $t > 0$ by the advection-diffusion equation
\begin{equation}
    \pdifft{c} + \nabla\cdot\JJ = 0 \quad \text{in } \Omega,
    \label{eq:adv_diff_strong}
\end{equation}
where the total concentration flux $\JJ$ is assumed to consist of an advective and a diffusive flux:
\begin{equation*}
    \JJ = c\uu - D\nabla c .
\end{equation*}
Here, $\uu$ is the CSF velocity field, governed by the Stokes equations~\Cref{eq:stokes_eqs},
and $D$ is the molecular diffusion coefficient of the solute.
Initially, we set $c(\xx, 0)=0$ for all $\xx\in\Omega$.
The ventricular walls are assumed to be impermeable (admitting \emph{no flux}: $\JJ\cdot\nn=0$),
except at the anterior/posterior boundary $\Gamma_p$.
At the anterior/posterior boundary, we permit the solute concentration to be advected into
and out of the domain in a manner that ensures conservation of mass
(see~\Cref{subsec:adv_diff_eq_weak_form} for details).

@@

%-----DISCRETIZATION AND MODEL IMPLEMENTATION-----%
\subsection{Numerical approximation of the Stokes equations}
The Stokes equations~\Cref{eq:stokes_eqs} are discretized with a finite element method
in space, while in time, we employ a first-order implicit Euler discretization.
We set the timestep size $\Delta t = 0.02252$ s such that 20 timesteps represent one cardiac cycle.
For the spatial discretization, we use piecewise linear Brezzi-Douglas-Marini (BDM) elements for the velocity.
The BDM elements have degrees of freedom associated with integral moments of the
velocity component normal to the facets of elements~\cite{Brezzi1985TwoProblems},
thus we enforce the impermeability condition $\uu\cdot\nn=0$ strongly.
We impose the boundary conditions for both the tangential traction $\bsigpar$ and
the normal pressure $\Tilde{p}(t)$ weakly.
For the pressure, we use piecewise constant (zeroth order discontinuous Galerkin (DG)) elements.
The resulting total number of degrees of freedom at each time step is 943 820.
The element combination $\mathrm{BDM}_1$--$\mathrm{DG}_{0}$ yields a non-conforming
discretization scheme for the Stokes equations~\cite{Stenberg1989SomeEquations}.
To ensure stability of the scheme, continuity in tangential velocities on interior facets
is weakly enforced with a penalty parameter $\gamma=10$~\cite{Hong2016AEquations}.
Since the divergence of the velocity is a subspace of the pressure space for this scheme,
the divergence-free condition of the velocity is satisfied exactly on each
element, ensuring mass conservation.~\cite{Boffi2008FiniteProblem}

@@

Constricting the connection between the anterior and middle ventricle leads to
slower transport to the anterior ventricle (ROI 5),
and a slightly higher mean concentration in ROIs 2--4 and 6, compared to the original geometry.
On the other hand, constricting the duct connecting the middle and the posterior ventricle
leads to less transport into the posterior ventricle (ROI 6),
and slightly more transport towards the anterior ventricle (ROI 5).
Faster transport to the anterior ventricle (ROI 5) and slower transport to the
posterior ventricle (ROI 6) is also the case for the geometry with a shrunk middle ventricle.
Additionally, for the latter configuration, the transport upstream of CSF flow towards the
ventral-posterior region of the middle ventricle (ROI 3) is slowed down.
For the mesh where all three ventricles are modified,
we observe higher retention of solute in the middle ventricle (ROIs 2--4),
while both the anterior and the posterior ventricles (ROIs 5 and 6) have lower mean concentrations.
Altogether, our results show that the geometry and width of the ventricular ducts matter
for solute transport. This may explain why we observed larger differences
in the cilia-deficient computational model compared to the experimental data in
\Cref{fig:fig4}, as the model and the fish have different duct diameters.

\lyng{HH: Quantify differences in terms of percentage changes in the
final mean concentration values, maybe with a bar plot?}
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure6_compare_modified_geometries.png}
    \caption{\textbf{a.} The original computational mesh of the ventricular geometry.
    \textbf{b.} The four modified versions of the computational geometry. Compared
    to the original mesh, the volumes of the deformed meshes are reduced by 3.9\%, 4.1\%, 1.4\% 
    and 9.4\%.
    \textbf{c.} Mean concentration $\cbar(t)$ simulated with the original and modified geometries
    in the 6 ROIs. 
    \mk{MK: How do the different geoemtries compare in terms of volume,
    areas where cilia are active?} \lyng{HH: Good question, will add this information.}}
    \label{fig:fig6}
\end{figure}

@@

\subsection{Computational model assumptions and boundary conditions}
To model CSF, we employed slip boundary conditions at the ventricular walls,
contrasting the classical choice of a noslip boundary condition.
The use of slip boundary conditions is motivated by experimental observations of the
flow field in zebrafish embryo brain ventricles~\cite{Olstad2019CiliaryDevelopment}.
It has been observed, both previously and in this work, that the fluid velocity increases
towards the wall~(\Crefpanels{fig:fig1}c).
Using slip boundary conditions allows for this behavior.
A couple of previous continuum model simulation studies of
cilia-mediated flow~\cite{Siyahhan2014FlowVentricles, Thouvenin2020OriginCanal}
have modeled cilia as body forces. We chose to model the cilia as a force exerted tangentially on the boundary.
First, we deem this a good representation of the real nature of cilia,
since they are in reality microscale structures confined to the ventricular walls that
generate flow through collective motion. Second, it avoids the need of excessive mesh
refinement near the ciliated boundaries in the case that a
body force were to be applied in a cilia layer of depth around
5 \textmu m~\cite{Salman2022ComputationalEmbryo}.

In our CSF flow model, the ciliated boundaries are homogeneous,
whereas in reality, cilia populate the ventricle walls in a
heterogeneous manner~\cite{Olstad2019CiliaryDevelopment}. Heterogeneous spatial
arrangement of cilia, both in location and beating direction,
has been shown to enhance particle clearance in airway cilia arrays
in mice~\cite{Ramirez-SanJuan2020Multi-scaleArrays}. Incorporating spatial arrangement heterogeneity
in our model could possibly result in different transport timescales,
but we would not expect notable changes to the general trends of the macroscopic
transport that we study here, since the effects of introducing heterogeneous spacing
of cilia would alter the magnitude of cilia forces, but not their direction.
On the other hand, introducing more heterogeneous beating directionality could
alter the transport timescales as well as the directionality. In our model, all of the applied cilia
tangential traction is aligned with the rostrocaudal axis, and more heterogeneity
in this alignment to reflect heterogeneity in the beating directionality of
cilia would presumably affect both magnitude and directionality of advective transport.
Previous studies have demonstrated that changes to beating directionality affects transport,
and that specific coordination of the ciliary beating is important in the brain ventricles for
flow and signaling processes~\cite{Eichele2020Cilia-drivenVentricle, Olstad2019CiliaryDevelopment,
Faubel2016Cilia-basedVentricles, Guirao2010CouplingCilia, Afzelius2004Cilia-relatedDiseases},
but also in other phenomena such as maintaining healthy functioning of mucous transport in
airways~\cite{Rayner1996CiliarySyndrome, Schneiter2021Multi-scaleFunction,
Bustamante-Marin2019LackClearance, Tsukita2012CoordinatedFeet} and in nodal flow which establishes
left-right symmetry in mammals~\cite{Yoshiba2014RolesSymmetry, Hirokawa2006NodalAsymmetry,
Sawamoto2006NewBrain}. Although these studies were mostly on spatial scales resolving individual cilia,
we could envision the same effects in our continuum model,
similar to the results of previous numerical studies~\cite{Ramirez-SanJuan2020Multi-scaleArrays,
Thouvenin2020OriginCanal, Yoshida2022EffectVentricles}.

We used an advection-diffusion equation to simulate protein transport in the brain ventricles.
Except for where we imposed a normal pressure boundary condition on the CSF flow,
we imposed a no-flux condition on the concentration. Imposing a pressure-driven
cardiac-related flow on certain parts of the CSF flow boundary leads to in- and outflow,
and the advective transport then requires handling of in- and outflux boundary conditions.
Reflecting our assumption that the ventricular system is closed, we set boundary conditions such
that the total mass is conserved in the limit that the timestep size is zero.
We handle the outflux by leaving the advective flux term as an unknown in the weak formulation
of the advection-diffusion equation. This has been referred to with different names in previous works,
names such as the ``free boundary condition"~\cite{Papanastasiou1992ACondition},
the ``no boundary condition"~\cite{Griffiths1997TheCondition}, and the 
``consistent flux boundary condition"~\cite{Lynch2020NumericalHemodynamics}. 
In leaving the advective flux term in the weak form we are lacking a boundary condition,
so the approach seems problematic in the continuous sense of the problem.
However, the discrete problem is well-posed, because the approach represents
constraining higher-order derivatives of the concentration~\cite{Griffiths1997TheCondition}.

On the inflow boundary, we applied an approximate periodic boundary condition by setting
the inflow boundary flux at timestep $n$ equal to the outflow boundary flux at the previous
timestep $n-1$. This is expected to be of low-order accuracy, and for high Péclet numbers,
the boundary condition affects the concentration field in the vicinity of the in- and outflow boundaries
due to large advective fluxes. For the simulation scenarios we considered, both in terms of model parameters
and time span, the photoconversion region where solute originates is far enough away from the in- and outflow
boundaries for the flux boundary condition to not significantly impact the results. Nevertheless,
we deliberately excluded the space proximal to in- and outflow boundaries in
the anterior and posterior regions of interest (5 and 6) to limit adverse effects.

@@

We heuristically determined the parameter $\tau = 0.65$ mPa, which determines
the magnitude of the cilia tangential traction boundary conditions. The maximum force applied
was in the dorsal region of the middle ventricle, where the posterior region had a tangential
traction vector magnitude of $\vert\btau\vert_{\mathrm{max}} = 2.5$ mPa applied to the
boundary. A couple of previous simulation studies of cilia-mediated flow in brain ventricles have
used a volumetric force density to represent the cilia. Siyahhan~\emph{et al.}~\cite{Siyahhan2014FlowVentricles}
used a force expression that increased linearly from the wall to a maximum force density of
$f_{\mathrm{max}} = 526$ N/$\mathrm{m^3}$ at a distance 15 \textmu m away from the wall,
which represented cilium length. If we assumed a cilia array of the same thickness
$l = 15$ \textmu m in our model, our maximum traction parameter $\vert\btau\vert_{\mathrm{max}}$
would correspond to a maximum force density of 
\begin{equation*}
    f_c = \frac{\vert\btau\vert_{\mathrm{max}}}{l}
        = \frac{\num{2.5e-3} \ \mathrm{Pa}}{15 \ \mu\mathrm{m}}
        = 169 \ \mathrm{N/m^3},
\end{equation*}
which smaller than $f_{\mathrm{max}}$, albeit the same order of magnitude.
The length 15 \textmu m is the a typical length of cilia in human brain ventricles
(in general the length is between 10--15 \textmu m~\cite{Afzelius2004Cilia-relatedDiseases}),
while in zebrafish they are closer to 5 \textmu m~\cite{Salman2022ComputationalEmbryo}.
With $l=5$ \textmu m, our traction parameter corresponds to a maximum force density of
$f_c=507$ N/$\mathrm{m^3}$, which is in good agreement with $f_{\mathrm{max}} = 526$ N/$\mathrm{m^3}$
used by Siyahhan~\emph{et al.}~\cite{Siyahhan2014FlowVentricles}.
We note that they based $f_{\mathrm{max}}$ on reconstructions of flow velocities in murine brain ventricles.

@@

\subsection{Numerical approximation and convergence rates}
The Brezzi-Douglas-Marini and discontinuous Galerkin discretization scheme
($\mathrm{BDM}_1$-$\mathrm{DG}_0$) that we employ for the Stokes equations
is computationally more expensive than a more conventional approach based on
e.g. Taylor-Hood discretization~\cite{Stenberg1990ErrorProblem}
($\mathbb{P}_2-\mathbb{P}_1$ elements) or lower-order enriched elements~\cite{Brezzi2011MixedMethods},
such as the mini-element ($\mathbb{P}_1-\mathbb{P}_1$ with a velocity stabilization).
Here, $\mathbb{P}_k$ is the set of continuous Lagrange polynomials of order $k$.
We opted for the BDM-DG scheme because of its exactly-mass-conserving property,
since numerical experiments proved that a divergence-free velocity field was
paramount to attain a stable discretization of the advection-diffusion equation
in advection-dominated regions. Using conventional Taylor-Hood elements for the Stokes equations,
stability of the advection-diffusion discretization either
required excessive mesh refinement and/or problem-specific tuning of
hyperparameters when employing a Streamline-Upwind/Petrov Galerkin stabilization
method~\cite{Franca1992StabilizedModel} instead of the current DG method.
Probing numerical values of the divergence of
the velocity when using Taylor-Hood elements for the fluid equations revealed that,
on the standard computational mesh, the $L^2$ norm of $\nabla\cdot\uu_h$ was
in the order of $10^{-6}$, while the
maximum value of the divergence of the velocity was of order unity. This is a result
of the fact that incompressibility is only achieved in a (global) weak sense when
employing Taylor-Hood elements. On a refined mesh with roughly six times as many
computational cells as the standard mesh, the numerical values of
$\normltwo{\nabla\cdot\uu_h}$ and $\normlinf{\nabla\cdot\uu_h}$ were still
of order unity. This may lead to the
hyperbolic advective term in the advection-diffusion equation accumulating errors
that grow unstable over time.

@@

We will consider the function spaces $\VV$ and $Q$ based on the
Sobolev spaces $\mathbf{H}^1(\mathrm{div}; \Omega)$ and $L^2(\Omega)$,
respectively, defined as 
\begin{align*}
    L^2(\Omega) &= \bigg\{p \ : \ \intO{p^2} < \infty\bigg\} \\
    \mathbf{H}^1(\mathrm{div}; \Omega)
    &= \bigg\{\uu \ : \ \uu\in\mathbf{L}^2(\Omega), \nabla\cdot\uu\in L^2(\Omega)\bigg\}.
\end{align*}
Bold-faced characters for a space denotes a $d$--dimensional vector space,
where $d$ is the spatial dimension of $\Omega$. For example,
$\mathbf{H}^1(\Omega) = [H^1(\Omega)]^3$ if $\Omega$ is three-dimensional.
We will consider $p\in Q=L^2_0(\Omega)$, with subscript 0 meaning
that $\intO{p} = 0$, and $\uu\in\VV$ with
\begin{equation*}
    \VV =  \bigg\{\vv \ : \ \vv\in\mathbf{H}^1(\mathrm{div}; \Omega),
    \vv\cdot\nn=0 \ \mathrm{on} \ \Gamma\setminus\Gamma_{\mathrm{io}}\bigg\}.
\end{equation*}
Here $\Gamma_{\mathrm{io}}$ denotes an in- or outflow boundary.

\mk{
You should be careful here. The H(div) functions do not have a
well defined gradient (and in turn $\beps$). However, for the derivations of
the continuous problem you don't need H(div). I would derive all the weak forms
in $V = \left\{ v\in H^1, v\cdot n = \text{on} \Gamma\right\}$. Then in A.2 you say we
chose spaces $V_h$ which are are not conforming, i.e. it does not hold that $V_h\subset V$.
Instead you have $V_h\subset H(div)$. Reasons is better mass csrv and as a
bonus the bcs can be enforced strongly. Note that H(div) functions can have
jumps in tangential components on surfaces in $\Omega$. Then introduce the discrete weak form.  
}
To derive a weak formulation of the Stokes equations~\Cref{eq:appendix_stokes_eqs},
we begin by multiplying~\Cref{eq:BDM_stokes_mom} with a
test function $\vv\in\VV$ and integrating over the domain:
\begin{equation*}
    \intO{\rho\pdifft{\uu}\cdot\vv} - \intO{(\nabla\cdot\bsig)\cdot\vv} = \mathbf{0}.
\end{equation*}
Performing integration by parts of the stress term and applying Green's first identity, we get
\begin{equation}
    \intO{\rho\pdifft{\uu}\cdot\vv}
    - \intG{\bsig\nn\cdot\vv}
    + \intO{2\mu\beps(\uu) : \beps(\vv)}
    - \intO{p\nabla\cdot\vv} = \mathbf{0}, \label{eq:modelA_weak_form1}
\end{equation}
where the symbol : signifies the matrix inner product,
$A:B=\text{trace}{(A^TB)}$ for matrices $A, B\in\mathbb{R}^{d\times d}$.

Further steps depend on the boundary conditions applied,
which is different for the three model versions we consider.
Before going through the different model versions, let us introduce
some notation used in defining slip boundary conditions. We make use of a decomposition
of any vector $\vv$ on the boundary $\Gamma$ into its normal and tangential components,
$\vv = \vv_{\perp} + \vv_{\parallel}$, where $\vv_{\perp}=(\vv\cdot\nn)\nn$ and
$\vv_{\parallel}=P_{\nn}(\vv)$. The tangential projection is defined as
\begin{equation}
    P_{\nn}(\vv) = (\mathbf{I} - \nn\otimes\nn)\vv,
    \label{eq:projection_operator_appendix}
\end{equation}
where $\nn$ is the outer unit normal of the surface. Introducing
$\hat{\bsig} = \bsig\nn$ for the traction (vector), the forcing due
to cilia is represented by constraining $\bsigpar$ on the part of the
boundary populated by cilia. In particular, we let
$\bsigpar = \btau(\xx, \rr) = \tau\lambda(\xx) P_{\nn}(\rr)$,
where the parameter $\tau=\num{6.5e-4}$ Pa was used to control the magnitude
of the tangential stresses, and chosen by calibrating simulated velocity
fields with experimental data~\cite{Olstad2019CiliaryDevelopment}.
The details of the function $\lambda(\xx)$ follow subsequent to deriving the weak form.

>>>>

\subsection{Computational model of solute transport after photoconversion}
We simulate transport of photoconverted proteins within the brain ventricles by
modeling transport of a solute with concentration $c(\xx, t)$ for $\xx\in\Omega$ and
time $t > 0$ by the advection-diffusion equation
\begin{equation}
    \pdifft{c} + \nabla\cdot\JJ = 0 \quad \text{in } \Omega,
    \label{eq:adv_diff_strong}
\end{equation}
where the total concentration flux $\JJ$ is assumed to consist of an advective and a diffusive flux:
\begin{equation*}
    \JJ = c\uu - D\nabla c .
\end{equation*}
Here, $\uu$ is the CSF velocity field, governed by the Stokes equations~\Cref{eq:stokes_eqs},
and $D$ is the molecular diffusion coefficient of the solute.
Initially, we set $c(\xx, 0)=0$ for all $\xx\in\Omega$.
The ventricular walls are assumed to be impermeable (admitting \emph{no flux}: $\JJ\cdot\nn=0$),
except at the anterior/posterior boundary $\Gp$.
At the anterior/posterior boundary, we permit the solute concentration to be advected into
and out of the domain in a manner that ensures conservation of mass
(see~\Cref{subsec:adv_diff_eq_weak_form} for details).

@@

%-----DISCRETIZATION AND MODEL IMPLEMENTATION-----%
\subsection{Numerical approximation of the Stokes equations}
The Stokes equations~\Cref{eq:stokes_eqs} are discretized with a finite element method
in space, while in time, we employ a first-order implicit Euler discretization.
We set the timestep size $\Delta t = 0.02252$ s such that 20 timesteps represent one cardiac cycle.
For the spatial discretization, we use piecewise linear Brezzi-Douglas-Marini (BDM)
elements for the velocity.~\cite{Brezzi1985TwoProblems, Nedelec1986AIR3}
The BDM elements have degrees of freedom associated with integral moments of the
velocity component normal to the facets of elements,
thus we enforce the impermeability condition $\uu\cdot\nn=0$ strongly.
We impose the boundary conditions for both the tangential traction $\bsigpar$ and
the normal pressure $\Tilde{p}(t)$ weakly.
For the pressure, we use piecewise constant (zeroth order discontinuous Galerkin (DG)) elements.
The resulting total number of degrees of freedom at each time step is 943 820.
The element combination $\mathrm{BDM}_1$--$\mathrm{DG}_{0}$ yields a non-conforming
discretization scheme for the Stokes equations~\cite{Stenberg1989SomeEquations}.
To ensure stability of the scheme, continuity in tangential velocities on interior facets
is weakly enforced with a penalty parameter $\gamma=10$~\cite{Hong2016AEquations}.
Since the divergence of the velocity space is a subspace of the pressure space for this scheme,
the divergence-free condition of the velocity is satisfied exactly on each
element, ensuring mass conservation.~\cite{Boffi2008FiniteProblem}

@@

Constricting the connection between the anterior and middle ventricle leads to
slower transport to the anterior ventricle (ROI 5),
and a slightly higher mean concentration in ROIs 2--4 and 6, compared to the original geometry.
On the other hand, constricting the duct connecting the middle and the posterior ventricle
leads to less transport into the posterior ventricle (ROI 6),
and slightly more transport towards the anterior ventricle (ROI 5).
Faster transport to the anterior ventricle (ROI 5) and slower transport to the
posterior ventricle (ROI 6) is also the case for the geometry with a shrunk middle ventricle.
Additionally, for the latter configuration, the transport upstream of CSF flow towards the
ventral-posterior region of the middle ventricle (ROI 3) is slowed down.
For the mesh where all three ventricles are modified,
we observe higher retention of solute in the middle ventricle (ROIs 2--4),
while both the anterior and the posterior ventricles (ROIs 5 and 6) have lower mean concentrations.
Altogether, our results show that the geometry and width of the ventricular ducts matter
for solute transport. This may explain why we observed larger differences
in the cilia-deficient computational model compared to the experimental data in
\Cref{fig:fig4}, as the model and the fish have different duct diameters.
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure6_compare_modified_geometries.png}
    \caption{\textbf{a.} The original computational mesh of the ventricular geometry.
    \textbf{b.} The four modified versions of the computational geometry. Compared
    to the original mesh, the volumes of the deformed meshes are reduced by 3.9\%, 4.1\%, 1.4\% 
    and 9.4\%.
    \textbf{c.} Mean concentration $\cbar(t)$ simulated with the original and modified geometries
    in the 6 ROIs. 
    \mk{MK: How do the different geoemtries compare in terms of volume,
    areas where cilia are active?} \lyng{HH: Good question, will add this information.}}
    \label{fig:fig6}
\end{figure}

@@

\subsection{Computational model assumptions and boundary conditions}
To model CSF, we employed slip boundary conditions at the ventricular walls,
contrasting the classical choice of a noslip boundary condition.
The use of slip boundary conditions is motivated by experimental observations of the
flow field in zebrafish embryo brain ventricles, both previously and in this work, in which
the fluid velocity increases towards the
wall~(\Crefpanels{fig:fig1}c)~\cite{Olstad2019CiliaryDevelopment}. Slip boundary
conditions allow for this behavior, in contrast to a noslip condition where the velocity
has to go to zero at the wall.

As opposed to a couple of previous continuum model simulation studies of
cilia-mediated flow where cilia were represented by body
forces,~\cite{Siyahhan2014FlowVentricles, Thouvenin2020OriginCanal}
we chose to model the cilia as a force
exerted tangentially on the boundary.
We deem this a good representation of the real nature of cilia,
since they are microscale structures confined to the ventricular walls that
generate flow through collective motion. Furthermore, it avoids the need of mesh
refinement near the ciliated boundaries in the case that a
body force were to be applied in a cilia layer of depth around
5 \textmu m~\cite{Salman2022ComputationalEmbryo}.

The ciliated boundaries in our CSF flow model are regionally homogeneous,
whereas in reality, cilia populate the ventricle walls in a
heterogeneous manner~\cite{Olstad2019CiliaryDevelopment}. Heterogeneous
arrangement of cilia, both in location, beating direction and beating frequency,
has been shown to enhance particle clearance in airway cilia arrays
in mice~\cite{Ramirez-SanJuan2020Multi-scaleArrays}.
Incorporating spatial arrangement heterogeneity in our model could possibly
result in different transport timescales,
but we would not expect notable changes to the general trends of the macroscopic
transport that we study here, since the effects of introducing heterogeneous spacing
of cilia would alter the magnitude of cilia forces, but not their direction.
On the other hand, introducing more heterogeneous beating directionality could
alter the transport timescales as well as the directionality.
In our model, all of the applied cilia
tangential traction is aligned with the rostrocaudal axis, and more heterogeneity
in this alignment to reflect heterogeneity in the beating directionality of
cilia would presumably affect both magnitude and directionality of advective transport.
Previous studies have demonstrated that changes to beating directionality affects transport,
and that specific coordination of the ciliary beating is important in the brain ventricles for
flow and signaling processes~\cite{Eichele2020Cilia-drivenVentricle, Olstad2019CiliaryDevelopment,
Faubel2016Cilia-basedVentricles, Guirao2010CouplingCilia, Afzelius2004Cilia-relatedDiseases},
but also in other phenomena such as maintaining healthy functioning of mucous transport in
airways~\cite{Rayner1996CiliarySyndrome, Schneiter2021Multi-scaleFunction,
Bustamante-Marin2019LackClearance, Tsukita2012CoordinatedFeet} and in nodal flow which establishes
left-right symmetry in mammals~\cite{Yoshiba2014RolesSymmetry, Hirokawa2006NodalAsymmetry,
Sawamoto2006NewBrain}.
Although these studies were mostly on spatial scales resolving individual cilia,
we could envision the same effects in our continuum model,
similar to the results of previous numerical studies~\cite{Ramirez-SanJuan2020Multi-scaleArrays,
Thouvenin2020OriginCanal, Yoshida2022EffectVentricles}.

To simulate protein transport in the brain ventricles, we numerically solved
an advection-diffusion equation. Except for where we imposed a normal pressure
boundary condition on the CSF flow,
we imposed a no-flux condition on the concentration. Imposing a pressure-driven
cardiac-related flow on certain parts of the CSF flow boundary leads to in- and outflow,
and the advective transport then requires handling of in- and outflux boundary conditions.
We set boundary conditions such that the total mass is conserved in the
limit that the timestep size is zero, reflecting our assumption that the
ventricular system is closed.
We handled the outflux by leaving the advective flux term as an unknown in the weak formulation
of the advection-diffusion equation. This has been referred to with different names
in previous works, names such as the ``free boundary condition"~\cite{Papanastasiou1992ACondition},
the ``no boundary condition"~\cite{Griffiths1997TheCondition}, and the 
``consistent flux boundary condition"~\cite{Lynch2020NumericalHemodynamics}. 
In leaving the advective flux term in the weak form we are lacking a boundary condition,
which is problematic in the continuous sense of the problem.
However, the discrete problem is well-posed, because the approach represents
constraining higher-order derivatives of the concentration~\cite{Griffiths1997TheCondition}.

Our approach can be interpreted as an approximate periodic boundary condition, where we set
the inflow boundary flux at timestep $n$ equal to the outflow boundary flux at the previous
timestep $n-1$. This is expected to be of low-order accuracy, and for high Péclet numbers,
the boundary condition affects the concentration field in the vicinity of the
in- and outflow boundaries
due to large advective fluxes. For the simulation scenarios we considered,
both in terms of model parameters
and time span, the photoconversion region where solute originates
is far enough away from the in- and outflow
boundaries for the flux boundary condition to not significantly impact the results. Nevertheless,
we deliberately excluded the space proximal to in- and outflow boundaries in
the anterior and posterior regions of interest (5 and 6) to limit adverse effects.

@@

We heuristically determined the parameter $\tau = 0.65$ mPa, which determines
the magnitude of the cilia tangential traction boundary conditions. The maximum force applied
was in the dorsal region of the middle ventricle, where the posterior region had a tangential
traction vector magnitude of $\vert\btau\vert_{\mathrm{max}} = 2.5$ mPa applied to the
boundary. Previous simulation studies of cilia-mediated flow in brain ventricles have
used a volumetric force density to represent the cilia.
Siyahhan~\emph{et al.}~\cite{Siyahhan2014FlowVentricles}
used a force expression that increased linearly from the wall to a maximum force density of
$f_{\mathrm{max}} = 526$ N/$\mathrm{m^3}$ at a distance 15 \textmu m away from the wall,
which represented cilium length. If we assumed a cilia array of the same thickness
$l = 15$ \textmu m in our model, our maximum traction parameter $\vert\btau\vert_{\mathrm{max}}$
would correspond to a maximum force density of 
\begin{equation*}
    f_c = \frac{\vert\btau\vert_{\mathrm{max}}}{l}
        = \frac{\num{2.5e-3} \ \mathrm{Pa}}{15 \ \mu\mathrm{m}}
        = 169 \ \mathrm{N/m^3},
\end{equation*}
which is smaller than $f_{\mathrm{max}}$, albeit the same order of magnitude.
The length 15 \textmu m is the a typical length of cilia in human brain ventricles
(in general the length is between 10--15 \textmu m~\cite{Afzelius2004Cilia-relatedDiseases}),
while in zebrafish they are closer to 5 \textmu m~\cite{Salman2022ComputationalEmbryo}.
With $l=5$ \textmu m, our traction parameter corresponds to a maximum force density of
$f_c=507$ N/$\mathrm{m^3}$, which is in good agreement with
$f_{\mathrm{max}} = 526$ N/$\mathrm{m^3}$
used by Siyahhan~\emph{et al.}~\cite{Siyahhan2014FlowVentricles}.
We note that they based $f_{\mathrm{max}}$ on reconstructions of
flow velocities in murine brain ventricles.

@@

\subsection{Numerical approximation and convergence rates}
The Brezzi-Douglas-Marini and discontinuous Galerkin discretization scheme
($\mathrm{BDM}_1$-$\mathrm{DG}_0$) that we employ for the Stokes equations
is computationally more expensive than a more conventional approach based on
e.g. Taylor-Hood discretization~\cite{Stenberg1990ErrorProblem}
($P_2-P_1$ elements) or lower-order enriched
elements~\cite{Brezzi2011MixedMethods},
such as the mini-element ($P_1-P_1$ with a velocity stabilization).
Here, $P_k$ is the space of continuous Lagrange polynomials of degree less
than or equal to $k$.
We opted for the BDM-DG scheme because of its exactly-mass-conserving property,
since numerical experiments proved that a divergence-free velocity field was
fundamental to attain a stable discretization of the advection-diffusion equation
in advection-dominated regions. Using conventional
Taylor-Hood elements for the Stokes equations
stability of the advection-diffusion discretization either
required excessive mesh refinement and/or problem-specific tuning of
hyperparameters when employing a Streamline-Upwind/Petrov Galerkin stabilization
method~\cite{Franca1992StabilizedModel}, which is why we also opted for a DG method
for the advection-diffusion equation.
Probing numerical values of the divergence of
the velocity when using Taylor-Hood elements for the fluid equations revealed that,
on the standard computational mesh, the $L^2$ norm of $\nabla\cdot\uu_h$ was
in the order of $10^{-6}$, while the
maximum value of the divergence of the velocity was of order unity. This is a result
of the fact that incompressibility is only achieved in a (global) weak sense when
employing Taylor-Hood elements. On a refined mesh with roughly six times as many
computational cells as the standard mesh, the numerical values of
$\normltwo{\nabla\cdot\uu_h}$ and $\normlinf{\nabla\cdot\uu_h}$ were still
of order unity. This may lead to the
hyperbolic advective term in the advection-diffusion equation accumulating errors
that grow unstable over time.

@@

To derive weak formulations of the continuous problem defined by the Stokes
equations~\Cref{eq:appendix_stokes_eqs}, we consider the function space
\begin{align*}
    \VV &= \bigg\{\vv \ : \ \vv\in\mathbf{H}^1(\Omega),
    \vv\cdot\nn=0 \ \mathrm{on} \ \Gamma\setminus\Gp\bigg\},
\end{align*}
for the velocity and $Q=L^2(\Omega)$ for the pressure.
Here $\Gp$ is the anterior/posterior boundary where we impose a normal pressure
boundary condition for the model versions that consider fluid motion induced by
cardiac pulsations. Bold-faced characters for a space denotes a $d$--dimensional vector space,
where $d$ is the spatial dimension of $\Omega$. For example,
$\mathbf{H}^1(\Omega) = [H^1(\Omega)]^3$ if $\Omega$ is three-dimensional.

To derive a weak formulation of the Stokes equations~\Cref{eq:appendix_stokes_eqs},
we begin by multiplying~\Cref{eq:BDM_stokes_mom} with a
test function $\vv\in\VV$ and integrating over the domain:
\begin{equation*}
    \intO{\rho\pdifft{\uu}\cdot\vv} - \intO{(\nabla\cdot\bsig)\cdot\vv} = \mathbf{0}.
\end{equation*}
Performing integration by parts of the stress term and applying Green's first identity, we get
\begin{equation}
    \intO{\rho\pdifft{\uu}\cdot\vv}
    - \intG{\bsig\nn\cdot\vv}
    + \intO{2\mu\beps(\uu) : \beps(\vv)}
    - \intO{p\nabla\cdot\vv} = \mathbf{0}, \label{eq:modelA_weak_form1}
\end{equation}
where the symbol : signifies the matrix inner product,
$A:B=\text{trace}{(A^TB)}$ for matrices $A, B\in\mathbb{R}^{d\times d}$.

Further steps depend on the boundary conditions applied,
which is different for the three model versions we consider in this paper.
Before proceeding, let us introduce some notation used in defining the
slip boundary conditions. We make use of a decomposition
of any vector $\vv$ on the boundary $\Gamma$ into its normal and tangential components,
$\vv = \vv_{\perp} + \vv_{\parallel}$, where $\vv_{\perp}=(\vv\cdot\nn)\nn$ and
$\vv_{\parallel}=P_{\nn}(\vv)$. The tangential projection is defined as
\begin{equation}
    P_{\nn}(\vv) = (\mathbf{I} - \nn\otimes\nn)\vv,
    \label{eq:projection_operator_appendix}
\end{equation}
where $\nn$ is the outer unit normal of the surface. Introducing
$\hat{\bsig} = \bsig\nn$ for the traction (vector), the forcing due
to cilia is represented by constraining $\bsigpar$ on the part of the
boundary populated by cilia. In particular, we let
$\bsigpar = \btau(\xx, \rr) = \tau\lambda(\xx) P_{\nn}(\rr)$,
where the parameter $\tau=\num{6.5e-4}$ Pa was used to control the magnitude
of the tangential stresses, and chosen by calibrating simulated velocity
fields with experimental data~\cite{Olstad2019CiliaryDevelopment}.
The details of the function $\lambda(\xx)$ follow subsequent to deriving the weak form.

We choose $\mathbf{r} = \rr^+ = (1, 0, 1)$ or $\mathbf{r} = \rr^- = -(1, 0, 1)$
to impose forces perpendicular to the $y$-axis.
The reason for this is that all of the experimental images we compare
with are taken in the $xz$-plane,
rendering flow features in the $y$-direction unobservable.
In the following, we consider the way of handling the stress
boundary integral for each model version successively.
